Production of software and data is not the purpose of this project
as such. However, considerable software will be developed as proof-of-concept
or in the course of doing simulations.
Issues of confidentiality are largely irrelevant to this project,
but we will give careful consideration to the matter
of \textbf{reproducible research}.
The lead PI will be responsible for execution of the plan.

Software, data, and publications will be released
under Open Source licenses.
The universities and investigators will retain all appropriate
intellectual property rights, including copyright and patent rights.

\heading{Preservation.} We will use distributed version control
(repositories) to manage the software and related material. This
results in complete backups at multiple sites; provenance that
includes a complete, reproducible record of development history of the
software; and robust mechanisms for coordinating simultaneous
development across multiple sites.

\heading{Analysis and simulation software}
In the process of exploring hardware redesign
we will, of necessity, rely on simulations.
All software developed here will be placed in a repository
and ultimately released, together with driver scripts
for all published results.

\heading{Proof-of-concept software}
The project will produce software to prove the viability
of the \ac{IMP} model as a basis of co-design. This software, while not
a full-fledged programming system, will serve as proof-of-concept
and is as such worth releasing. We will use a software repository
that on publication of our results will be opened to the public.
Build scripts and drivers for test cases will be provided.

\heading{Publications and presentation}

This includes research papers and similar products accepted for
publication or presentation. We will provide full access to all such
products within ninety days of acceptance. We will also provide full
access to education and outreach materials (e.g.\ presentation slides
that do not contain programmatic information), even when used only in
informal settings.

The final report will include an
inventory of products managed under this plan, together with relevant
information such as availability and status.
